---
title: "An introduction to SigMoS package"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SigMoS-vignette}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEngine{rmarkdown::render}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Use the following commands to install the SigMoS package from GitHub:

```{r setup}
library(devtools)
#install_github("MartaPelizzola/SigMoS")
library(ggplot2)
library(SigMoS)
```

A typical mutational count data set we considered for the SigMoS package has 96 rows (equal to the number of mutation types with 6 base mutations, when assuming strand symmetry and the 4 flanking nucleotides on each side) and as many columns as many patients are included in the data set. 

We show below the different options available in this R package. 

## 1. SigMoS with Poisson-NMF
When choosing the Poisson distributional assumption we can apply the model selection function with the Poisson model as follows:
```{r poisson}
set.seed(1234567)
k_vec <- 2:8
nsig_pois <- list()
cost <- c()
for (i in k_vec){
  nsig_pois[[i-1]] = sigmos(data = BRCA21, k = i, method = "Poisson")
  cost = c(cost, nsig_pois[[i-1]]$cost_k)
}
k_poisson = k_vec[which.min(cost)]
cost # cost for the different values of the number of signatures
k_poisson # best number of signatures for this dataset under the Poisson model
```

=======

## 2. SigMoS with NB-NMF
Following the same procedure as for the Poisson case, we can also choose the Negative Binomial distributional assumption and find the optimal number of signatures under the Negative Binomial model with our model selection approach.


```{r NBpatientspecific, warning = FALSE}
set.seed(123456)
k_vec <- 2:8
nsig_nb_ps <- list()
cost <- c()
for (i in k_vec){
  nsig_nb_ps[[i-1]] = sigmos(data = BRCA21, k = i, method = "NB", 
                             patient_specific = TRUE)
  cost <- c(cost, nsig_nb_ps[[i-1]]$cost_k)
}
k_NB_ps <- k_vec[which.min(cost)]
cost # cost for the different values of the number of signatures
k_NB_ps # best number of signatures for this dataset under the Poisson model
```

## 3. Residual analysis as in Pelizzola et al. 2022
## 3a. Residual analysis under the Poisson distributional assumption
Given these results we can check the residuals and obtain the following residual plot:

```{r poissonRSD, warning = FALSE}
res_pois = sigmos(data = BRCA21, k = 3, method = "Poisson")
W_p <- res_pois$Exposures
H_p <- res_pois$Signatures
rsd_p = (BRCA21 - W_p%*%H_p)
rsd_norm = rsd_p/(sqrt(W_p%*%H_p))

funupper = function(x) 2*sqrt(x)
funlower = function(x) -2*sqrt(x)

dataPois = data.frame(obs = as.vector(W_p%*%H_p),Residuals = as.vector(rsd_p), 
                      residual_norm = as.vector(rsd_norm))

ggplot(dataPois, aes(x = obs, y = Residuals))+
  geom_point()+
  scale_x_log10()+
  stat_function(fun = funupper, color = "#0072B2", size = 1 ) + 
  stat_function(fun = funlower, color = "#0072B2", size = 1 ) +
  ylim(-100,100) + 
  theme(strip.background = element_blank(), strip.text = element_blank(),
        axis.text = element_text(size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        line = element_line(margin(b = 6, l = 6)), 
        legend.position = "none", 
        plot.title = element_text(size = 20)) +
  xlab("Estimated mean") + ylab("Raw residuals")

```

The plot shows overdispersion under the Poisson model, thus we proceed considering the negative binomial model. 

## 3b. Residual analysis under the Negative Binomial distributional assumption

```{r NBresiduals, warning = FALSE}
alpha_ps <- alphaNR(data = BRCA21, k = 3, patient_specific = TRUE)
res_nb_ps = sigmos(data = BRCA21, k = 3, method = "NB", patient_specific = TRUE)

W_nb_ps <- res_nb_ps$Exposures
H_nb_ps <- res_nb_ps$Signatures
rsd_nb_ps = (BRCA21 - W_nb_ps%*%H_nb_ps)
rsd_norm = rsd_nb_ps/(sqrt(W_nb_ps%*%H_nb_ps))

funuppermin = function(x) 2*sqrt(x*(1+x/min(alpha_ps)))
funlowermin = function(x) -2*sqrt(x*(1+x/min(alpha_ps)))
funuppermax = function(x) 2*sqrt(x*(1+x/max(alpha_ps)))
funlowermax = function(x) -2*sqrt(x*(1+x/max(alpha_ps)))
funuppermedian = function(x) 2*sqrt(x*(1+x/median(alpha_ps)))
funlowermedian = function(x) -2*sqrt(x*(1+x/median(alpha_ps)))

dataNB = data.frame(obs = as.vector(W_nb_ps%*%H_nb_ps),
                    Residuals = as.vector(rsd_nb_ps), 
                    residual_norm = as.vector(rsd_norm))

ggplot(dataNB, aes(x = obs, y = Residuals))+
  geom_point()+
  scale_x_log10()+
  stat_function(fun = funuppermin, color = "#009E73", size = 1, alpha = 0.5) + 
  stat_function(fun = funlowermin, color = "#009E73", size = 1, alpha = 0.5) +
  stat_function(fun = funuppermax, color = "#009E73", size = 1, alpha = 0.5 ) + 
  stat_function(fun = funlowermax, color = "#009E73", size = 1, alpha = 0.5 ) +
  stat_function(fun = funuppermedian, color = "#009E73", size = 1 ) + 
  stat_function(fun = funlowermedian, color = "#009E73", size = 1 ) +
  ylim(-100,100) + 
  theme(strip.background = element_blank(), strip.text = element_blank(),
        axis.text = element_text(size=14),
        axis.title.x = element_text(size=20),
        axis.title.y = element_text(size=20),
        line = element_line(margin(b = 6, l = 6)), 
        legend.position = "none", 
        plot.title = element_text(size = 20)) +
  xlab("Estimated mean") + ylab("Raw residuals")
```

The plot shows that the Negative Binomial model provides a much better fit to the data. 

## 4. Comparison of the mutational signatures 

We provide a function to compare the estimated mutational signatures based on the cosine similarity:

```{r corsig, warning = FALSE}
?corSig
corSig(res_nb_ps$Signatures, res_pois$Signatures)
```

This function provides the full distance matrix between the signatures based on the cosine similarity and a vector of indexes to reorder the second matrix to match the first one. 

A comparison between the estimated signatures can also be of interest when trying to find signatures from the COSMIC database. The signatures from the COSMIC can be downloaded from:

```{r cosmic, warning = FALSE}
library(mutSignatures)
cosmic <- getCosmicSignatures(forceUseMirror = FALSE, asMutSign = TRUE)
```

We refer to this package (function msigPlot) or other packages such as SigProfilerExtractor (Islam et al. 2022) and sigfit (Gori et al. 2018) to plot the mutational signatures. 
