---
title: "SigMoS-vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SigMoS-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(SigMoS)

```

We provide a detailed description of the pipeline for analyzing mutational counts data. 

A typical mutational count data set we considered for the SigMoS package has 96 rows (equal to the number of mutation types when considering the 6 base mutations, when assuming strand symmetry and the 4 flanking nucleotides on each side) and as many columns as many patients are included in the data set. 

## Assuming poisson

```{r}
res_pois = sigmos(data = BRCA21, k = 3, method = "Poisson")
```


## Assuming negative binomial

```{r}
res_nb = sigmos(data = BRCA21, k = 3, method = "NB")
```

## Assuming negative binomial with patient specific dispersion

```{r}
res_nb = sigmos(data = BRCA21, k = 3, method = "NB", patient_specific=TRUE)
```
